\bf Teaching analysis reproducibility with accessibility: Enriching secondary documentation to support skill level diversity of users.

### 2022-02-11 jsdse/Manuscript.tex
elaborating on accessibility types

\bigskip
\begin{abstract}
Research reproducibility is essential for scientific development. Yet, rates of reproducibility
are low, especially in the natural sciences. As increasingly more research is relying on
computing tools and software, efforts for improving
reproducibility rates have focused on providing availability of research workflows
as computer code, as well as raw and processed data in a digital form.
However, making research products digitally available, does not
necessarily ensures that
they can be easily reached by the majority of the target user base, let alone the general population,
making them effectively inaccessible, counteracting availability, and hindering reproducibility.

In this paper, we argue that accessibility of resources is a key aspect for
successful research reproducibility.
While accessibility is a broad term that encompasses various aspects
of the human experience (visual, motor, auditory, cognitive, social, psychological,
educational, etc.), we particularly address barriers to accessibility at the
technological, social and pedagogical level.
Using an example within evolutionary biology, we identify barriers to
accessibility in the natural sciences:
%% list the barriers
specialized jargon, user programming experience, cognitive load,

@@

We recognize that accessibility is a term that encompasses various aspects
of the human experience (visual, motor, auditory, cognitive, social, psychological,
educational, etc.).
However, some forms of accessibility do not support reproducibility. For example,
Graphical User Interfaces
and
we focus on X types of accessibility

@@

      What we realize from these definitions is that in practice, something might
      be available but not necessarily accessible,
      hindering the realization of availability.
      To illustrate the practical difference between availability and accessibility we
      will use an allegory that goes like this:

@@

@@

To elaborate on our thesis,
%% we use a common scientific workflow in phylogenetics,
%% ** Jessica mentions that workflow is confusing as she thought it referred to the OpenTree
%% workflow, but we actually mean the more general workflow of reconstructing a phylogenetic tree.
%% She suggests something more general:
we choose an example from the research field of phylogenetics,
a discipline within evolutionary biology. Phylogenetics focuses
on investigating the history of shared ancestry of living organisms from biological data,
and represents
this evolutionary history in a diagram known as phylogeny (Figure \ref{fig:tree}).
Phylogenies provide the
basis to study and understand all biological processes in an evolutionary context
\citep{dobzhansky1973nothing}. Hence,
% Given the importance of phylogenies for biological research,
it appears that improving reproducibility rates
in phylogenetics has the potential to positively impact research across the natural sciences.

@@

The Open Tree of Life project (OpenTree) has developed a computational platform that provides
availability of results from vetted, state-of-the-art research in phylogenetics
\citep{opentreeoflife2019synth}.
%% by creating a standardized database of phylogenetic data that is used to synthesize
%% a single phylogenetic tree encompassing all life
OpenTree's database, the Phylesystem \citep{mctavish2015phylesystem}, provides
accessibility of phylogenetic data by standardizing organism names across biological databases,
and supporting digital data obtention in a computer readable format.

@@

is programmatically
available free of cost to users through OpenTree's Application Programming Interface
(API) services \citep{opentreeAPIs}.
Additionally, software packages written with open source, free of cost programming
languages commonly used in scientific research such as R and Python
\citep{baker2017scientific} have been developed as wrappers that have increased the availability
of OpenTree's API services, by making the OpenTree API functionalities more accessible
to a wider user audience \citep{michonneau2016rotl, mctavish2021opentree}.
Yet, the R and Python OpenTree API wrapper software packages have been mainly used
by computer-literate individuals
to seamlessly establish reproducible workflows to use and reuse expert phylogenetic
knowledge for biological research \citep{sanchez2019datelife, sanchez2021physcraper}
and education \citep{nguyen2020phylotastic, phylotasticedtools, galacticedtools}.
In the 5 years since its release, the R package wrapper for OpenTree \texttt{rotl}
has been cited by more than 190 biological research papers addressing various
topics \citep{googlerotl}.
Comparatively, other R packages for phylogenetics, such as \texttt{GGTREE}
and \texttt{mixOmics}, which were released one year after \texttt{rotl} have been
cited over a thousand times, which might indicate a delay \texttt{rotl}
adoption. Of course one potential explanation is that OpenTree services do not have
as wide an application as services provided by the afore mentioned R packages.
However, the
expressed enthusiasm (interest to use/inability to use the resource) by the target
 audience and larger community might point to a different issue.
%% We need a proper statistical distribution of citation numbers per year, to see
%% if rotl is an outlier or not.

@@

A practice that has become more and more widespread in programming-language pedagogical practices
is the use of typos and mistakes to normalize them for learners, and show them how
to solve them when they are outside the classroom \citep{shannon2015live}.
Yet, this is rarely done on written pedagogical materials.
Primary documentation focuses on demonstrating usage function with examples that
work seamlessly, without errors. We argue that the opposite is needed to support
adoption of reproducible workflows and support long term independence in learners
\citep{gaspar2007restoring}.
We demonstrate examples that do not work
as expected and exemplify ways to address them (Figure \ref{fig:error
}).

@@

We published the tutorials on a public, free license, free of cost, and free for
use and reuse repository and persistent website \citep{RopentreeTutorials, RopentreeTutorialsWebsite}.
The tutorial is available for the users to go back to any time they need it,
and to be passed on to other users (Figure \ref{fig:schedule}).

@@

We created a main version of the tutorial that is stable. Any updates to the tutorial are
published as new versions, or tutorials for new workshops \citep{CARPENTRIES}
 Versions presented at workshops are a copy from the original repository.
They represent a temporally stable snapshot of functions and workflows presented
during a workshop.

>>>>

\bigskip
\begin{abstract}
Research reproducibility is essential for scientific development. Yet, rates of reproducibility
are low, especially in the natural sciences. As increasingly more research is relying on
computing tools and software, efforts for improving
reproducibility rates have focused on digital availability of research workflows
as computer code, as well as raw and processed data in computer readable form.
However, making research products digitally available does not
necessarily ensures that
they can be easily reached by the majority of a target user base, let alone the general population,
making them effectively inaccessible, counteracting availability, and hindering reproducibility.

In this paper, we argue that accessibility of research products is a key aspect for
improving reproducibility rates.
While accessibility typically addresses differences in abilities (visual, hearing, cognitive).

across individuals, we particularly address barriers to accessibility at the
technological, social and pedagogical level.

Using an example within evolutionary biology, we identify barriers to
accessibility in the natural sciences:
%% list the barriers
specialized jargon, user programming experience, cognitive load,

@@

We recognize that the term accessibility is used in relation to various aspects of
the human experience. In relation to digital resources, these aspects have been broadly
categorized in four axis: perception, operability, understanding and robustness \citep{chisholm2001web, caldwell2008web}
% (visual, motor, auditory, cognitive, social, psychological, educational, etc.).
Some forms of accessibility do not support automatization nor scalability.
For example, Graphical User Interfaces (GUIs) are accessible, but not scalable and rarely automatizable.

we focus on X types of accessibility

@@

      What we realize from these definitions is that, in practice, something might
      be available but not necessarily accessible,
      hindering the realization of availability.
      To illustrate the practical difference between availability and accessibility we
      will use an allegory that goes like this:

@@

      ** Gauging accessibility of a resource is tricky because it depends on an
      individual experience of what is in effect \textbf{easy} for someone.
      \textbf{Easy} is a subjective quality that fully depends on individual perceptions
      resulting from individual contexts.
      As a society we are moving towards validating and welcoming all human experiences.
      To improve accessibility levels in the natural sciences, it
      is important to account for the factors that are creating different experiences
      of what is \textbf{easy}.

@@

To elaborate on our thesis,
%% we use a common scientific workflow in phylogenetics,
%% ** Jessica mentions that workflow is confusing as she thought it referred to the OpenTree
%% workflow, but we actually mean the more general workflow of reconstructing a phylogenetic tree.
%% She suggests something more general:
we choose an example from the research field of phylogenetics,
a discipline within evolutionary biology. Phylogenetics focuses
on investigating the history of shared ancestry of living organisms from biological data,
and represents
this evolutionary history in a diagram known as a phylogeny (Figure \ref{fig:tree}).
Phylogenies provide the
basis to study and understand all biological processes in an evolutionary context
\citep{dobzhansky1973nothing}. Hence,
% Given the importance of phylogenies for biological research,
it appears that improving reproducibility rates
in phylogenetics has the potential to positively impact research across the natural sciences.

@@

The Open Tree of Life project (OpenTree) has developed a computational platform that provides
digital availability of phylogenetic results from vetted, state-of-the-art research
in the field \citep{opentreeoflife2019synth}.
%% by creating a standardized database of phylogenetic data that is used to synthesize
%% a single phylogenetic tree encompassing all life
OpenTree's database, the Phylesystem \citep{mctavish2015phylesystem}, provides
usability of phylogenetic data by standardizing organism names across biological databases,
and supporting digital data obtention in a computer readable format.

@@

is programmatically
available free of cost to users through OpenTree's Application Programming Interface
(API) services \citep{opentreeAPIv3}.
Additionally, software packages written with open source, free of cost programming
languages commonly used in scientific research such as R and Python
\citep{baker2017scientific} have been developed as wrappers that have increased the availability
of OpenTree's API services, by making the OpenTree API functionalities more accessible
to a wider user audience \citep{michonneau2016rotl, mctavish2021opentree}.
Yet, the R and Python OpenTree API wrapper software packages have been mainly used
by computer-literate individuals
to seamlessly establish reproducible workflows to use and reuse expert phylogenetic
knowledge for biological research \citep{sanchez2019datelife, sanchez2021physcraper}
and education \citep{nguyen2020phylotastic, phylotasticedtools, galacticedtools}.
In the 5 years since its release, the R package wrapper for OpenTree \texttt{rotl}
has been cited by more than 190 biological research papers addressing various
topics \citep{googlerotl}.
Comparatively, other R packages for phylogenetics, such as \texttt{GGTREE}
and \texttt{mixOmics}, which were released one year after \texttt{rotl} have been
cited over a thousand times, which might indicate a delay \texttt{rotl}
adoption. Of course one potential explanation is that OpenTree services do not have
as wide an application as services provided by the afore mentioned R packages.
However, the
expressed enthusiasm (interest to use/inability to use the resource) by the target
 audience and larger community might point to a different issue.
%% We need a proper statistical distribution of citation numbers per year, to see
%% if rotl is an outlier or not.

@@

A practice that has become more and more widespread in programming-language pedagogical practices
is the use of typos and mistakes to normalize them for learners, and show them how
to solve them when they are outside the classroom \citep{shannon2015live}.
Yet, this is rarely done on written pedagogical materials.
Primary documentation focuses on demonstrating usage function with examples that
work seamlessly, without errors. We argue that the opposite is needed to support
adoption of reproducible workflows and support long term independence in learners
\citep{gaspar2007restoring}.
We demonstrate examples that do not work
as expected and exemplify ways to address them (Figure \ref{fig:error}).

@@

We published the tutorials on a public, free license, free of cost, and free for
use and reuse repository and persistent website \citep{RopentreeTutorials}.
The tutorial is available for the users to go back to any time they need it,
and to be passed on to other users (Figure \ref{fig:schedule}).

@@

We created a main version of the tutorial that is stable. Any updates to the tutorial are
published as new versions, or tutorials for new workshops \citep{wilson2006swc, SWCwebsite}
 Versions presented at workshops are a copy from the original repository.
They represent a temporally stable snapshot of functions and workflows presented
during a workshop.
